Plume-SPH 1.0: A 3D dust-gas volcano plume model based on Smoothed Particle Hydrodynamics

### 2017-02-18 Manuscripts_GMD.tex
Language revised

%% Copernicus Publications Manuscript Preparation Template for LaTeX Submissions
%% ---------------------------------
%% This template should be used for copernicus.cls
%% The class file and some style files are bundled in the Copernicus Latex Package, which can be downloaded from the different journal webpages.
%% For further assistance please contact Copernicus Publications at: production@copernicus.org
%% http://publications.copernicus.org/for_authors/manuscript_preparation.html
%% Please use the following documentclass and journal abbreviations for discussion papers and final revised papers.
%% 2-column papers and discussion papers
\documentclass[journal abbreviation, manuscript]{copernicus}
%% Journal abbreviations (please use the same for discussion papers and final revised papers)
% Archives Animal Breeding (aab)
% Atmospheric Chemistry and Physics (acp)
% Advances in Geosciences (adgeo)
% Advances in Statistical Climatology, Meteorology and Oceanography (ascmo)
% Annales Geophysicae (angeo)
% ASTRA Proceedings (ap)
% Atmospheric Measurement Techniques (amt)
% Advances in Radio Science (ars)
% Advances in Science and Research (asr)
% Biogeosciences (bg)
% Climate of the Past (cp)
% Drinking Water Engineering and Science (dwes)
% Earth System Dynamics (esd)
% Earth Surface Dynamics (esurf)
% Earth System Science Data (essd)
% Fossil Record (fr)
% Geographica Helvetica (gh)
% Geoscientific Instrumentation, Methods and Data Systems (gi)
% Geoscientific Model Development (gmd)
% Geothermal Energy Science (gtes)
% Hydrology and Earth System Sciences (hess)
% History of Geo- and Space Sciences (hgss)
% Journal of Sensors and Sensor Systems (jsss)
% Mechanical Sciences (ms)
% Natural Hazards and Earth System Sciences (nhess)
% Nonlinear Processes in Geophysics (npg)
% Ocean Science (os)
% Proceedings of the International Association of Hydrological Sciences (piahs)
% Primate Biology (pb)
% Scientific Drilling (sd)
% SOIL (soil)
% Solid Earth (se)
% The Cryosphere (tc)
% Web Ecology (we)
% Wind Energy Science (wes)
%% \usepackage commands included in the copernicus.cls:
%\usepackage[german, english]{babel}
%\usepackage{tabularx}
%\usepackage{cancel}
%\usepackage{multirow}
%\usepackage{supertabular}
%\usepackage{algorithmic}
%\usepackage{algorithm}
%\usepackage{amsthm}
%\usepackage{float}
%\usepackage{subfig}
%\usepackage{rotating}
\begin{document}
\title{Plume-SPH 1.0: A 3D dust-gas volcano plume model based on smoothed particle hydrodynamics}
% \Author[affil]{given_name}{surname}
\Author[1]{Zhixuan}{Cao}
\Author[1]{Abani}{Patra}
\Author[2]{Marcus}{Bursik}
\Author[3]{E. Bruce}{Pitman}
\Author[4]{Matthew}{Jones}
\affil[1]{Department of Mechanical and Aerospace Engineering, University at Buffalo, SUNY, New York, United State}
\affil[2]{Department of Geology, University at Buffalo, SUNY, New York, United State}
\affil[3]{Department of Mathematics, University at Buffalo, SUNY, New York, United State}
\affil[4]{Center for Computational Research,
University at Buffalo, SUNY, New York, United State}
%% The [] brackets identify the author with the corresponding affiliation. 1, 2, 3, etc. should be inserted.
\runningtitle{Plume-SPH 1.0}
\runningauthor{Z. Cao et al}
\correspondence{Abani Patra (abani.patra@gmail.com)}
\received{}
\pubdiscuss{} %% only important for two-stage journals
\revised{}
\accepted{}
\published{}
%% These dates will be inserted by Copernicus Publications during the typesetting process.
\firstpage{1}
\maketitle
\begin{abstract}
Volcanic ash, if persists in the atmosphere, will pose great threat on aircrafts and communities. VATDs (Volcanic ash transport and dispersion models) usually take output of plume models as source terms. The accuracy of source terms is crucial for VATDs. All of existing 3D (three dimensional) plume models are using mesh based methods. SPH (Smoothed particle hydrodynamics), as a mesh free method has several advantages over mesh based methods in modeling of multiphase free boundary flow. As an initial effort on exploiting the feasibility and advantage of SPH in volcano plume modeling we adopt a relative simpler model, a 3D dusty-gas dynamic model, targeting at capturing the more basic features of volcano plume. In this model, erupted material is assumed to be well mixed and represented by one phase while air is another phase. Immediate dynamic and thermal dynamic equilibrium between air and erupted material are assumed. Effect of wind is not considered. Several newly developed techniques in SPH are adopted and trimmed to address numerical challenges in simulating multiphase compressible turbulent flow with classical SPH method. The $SPH-\varepsilon$ turbulence model is adopted to capture sub-particle mixing. Heat exchange due to turbulence is then calculated with Reynolds analogy. Corrected formulism of SPH is used to handle tensile instability and deficiency of particle distribution near the boundaries. We also propose feasible ways to impose velocity inlet and pressure outlet boundary conditions, both of which are scarce in traditional implementations of SPH. The core solver of our model is parallelized with MPI (message passing interface) obtaining good weak scalability and strong scalability. The model is first verified by comparing velocity and concentration distribution along axis and cross section with experimental results of JPUE (jet or plume which is ejected from a nozzle into a uniform environment). The top height of Pinatubo eruption (15 June 1991) simulated by our model is consistent with both observation and existing 3D plume models. Profiles of several integrated variables are compared with these by existing 3D plume models and further verifies the correctness of our model. Analysis on the plume evolution process illustrated that this model is able to reproduce the basic physics of plume development. The comparison also implies that turbulence model plays a significant role in 3D volcano plume modeling.
\end{abstract}
\introduction  %% \introduction[modified heading if necessary]
\subsection{Volcanic ash hazards}
The dynamics of explosive eruption clouds has been the central issue of volcanology science for a long time \citep{settle1978volcanic, wilson1978control}.
The basic physics of volcanic plume has been understood for more than half a century \citep{morton1956turbulent, settle1978volcanic, wilson1978control}. The effects of magma type and vent conditions \citep{woods1988fluid, woods1995decompression}, atmospheric conditions \citep{ woods1993moist, sparks1997volcanic, bursik2001effect}, external surface water \citep{koyaguchi1996formation}, thermal disequilibrium, and particle fallout \citep{woods1991particle} have been assessed in increasing details based on observation, experiments and numerical studies.

@@

Hazards associated with explosive grey volcanoes include pyroclastic density currents (flows and surges), the widespread deposition of airfall tephra, and the threats posed by volcanic ash. 

@@

@@

Simulation of all possible hazards with one model is difficult due to diversity among these hazards. Our concern in current work is the risks that volcanic ash poses on aircrafts and communities. During volcanic eruptions, VATDs (volcanic ash transport and dispersion models) are used to forecast the location and movement of ash clouds over hours to days. Those VATDs use eruption source parameters, such as plume height, mass eruption rate, duration, and the mass fraction distribution of erupted debris finer than about $4 \Phi$ (or $63  \mu$m), which can remain in the cloud for many hours or days. Observational data for such parameters are usually unavailable in the first minutes or hours after an eruption is detected. Moreover, these input parameters are subject to change during an eruption, requiring rapid re-assignment of new parameters. Usually, plume models are used to provide source terms for VATDs. 

@@

Several 1D (one dimensional) volcano plume models have been developed in the past few decades. Based on the most basic 1D model \citep{woods1988fluid} which only accounts for basis conservation, recently developed 1D models (eg. \citep{bursik2001effect, mastin2007user, degruyter2012improving, woodhouse2013interaction, devenish2013using, de2015plume, folch2016fplume, pouget2016sensitivity}) tend to account for more physics effects. FPLUME-1.0 \citep{folch2016fplume} accounts for wind effect, entrainment of moisture, water phase change, particle fallout and re-entrainment and even wet aggregation of ash. However, In these 1D models, the entrainment of air is evaluated based on two coefficients: entrainment coefficient for the vertical plume and the entrainment coefficient that describes the effect of wind. Different 1D models adopt different entrainment coefficients based on specific formulation or calibration against well-documented case studies. The feedback from plume to atmosphere is also ignored in 1D models. Even though determination of essential parameters such as the entrainment is not based on first principles, such simple models nevertheless allow us to investigate the importance of physical mechanisms in a volcanic plume. In addition, these simplified models requires little computational resource and can run on standard personal computers or on web sites in very short time. As a result, 1D software for volcano plume development (such as \citep{267, 1194, 3541}), combined with VATDs (such as \citep{114, draxler2015hysplit}) are widely used in research and practice. While these 1D models can generate well-matched results with 3D (three dimensional) models for weak plumes, much greater variability is observed for strong plume scenarios, especially for local variables \citep{costa2016results}. In addition, the need for greater skills in associated hazards forecasts with 1D models requires 2D (two dimensional) and 3D models.\\
The development of 2D and 3D, time-dependent, and multiphase numerical models for volcano plume has provided new explanations for many features of explosive volcanism. One of these 3D models is PDAC (Pyroclastic Dispersion Analysis Code) \citep{neri2003multiparticle}  which is a non-equilibrium, multiphase, 3D compressible flow model. Conservation equations for each phase are solved separately with finite volume method. And a parallel computing version of PDAC was developed lately \citep{ongaro2007parallel}. Advanced numerical techniques, such as second order scheme and semi-explicit time upgrading, was also adopted afterwards to improve the accuracy of PDAC \citep{carcano2013semi}. 
Another 3D model, SK-3D \citep{suzuki2005numerical} is a 3D time-dependent fluid dynamics model that can reproduce the entrainment process of eruption clouds with relative simpler physics but high order of accuracy and high spatial resolution, which is applicable to time-dependent phenomena in actual volcanological simulations. 
A serials of simulations based on SK-3D was conducted, including determination of the entrainment coefficients of eruption columns as a function of height \citep{suzuki2010numerical}, study the effect of wind field on entrainment coefficient \citep{suzuki20133d}, establishment of the relationship between the observable quantities of the eruption clouds and the eruption conditions at the vent \citep{suzuki2009three} (previously, such relationship is based on 1D models and greatly depends on empirical constants), investigation of the effect of the intensity of turbulence in the umbrella cloud on dispersion and sedimentation of tephra \citep{koyaguchi2009effect}. 
While the SK-3D focuses on accurately capturing the entrainment caused by turbulent mixture with higher resolution and numerical method of higher order, PADC takes the disequilibrium between different phases into account and hence is a true multiphase model. Another 3D model, ATHAM ( Active Tracer High-Resolution Atmospheric Model) \citep{oberhuber1998volcanic} focuses more on microphysics within the plume. As pyroclastic flow is not the initial concern of ATHAM, immediately dynamics and thermodynamics equilibrium is assumed in ATHAM. The dynamic core of ATHAM solves the compressible Euler equations for momentum, pressure and temperature of the gas particle mixture \citep{oberhuber1998volcanic}. The subgrid-scale turbulence closure scheme differentiates between the horizontal and vertical directions \citep{herzog2003prognostic} to capture turbulent mixing. The cloud microphysics predicts the mass of hydrometeors in liquid and ice phase \citep{herzog1998effect}. Additional modules include radiation \citep{langmann1997radiative}, gas phase chemistry \citep{trentmann2002simulation} and gas scavenging by hydrometeors \citep{textor2003injection} were added lately. A further extension was made to include particle aggregation in 2006 \citep{textor2006volcanic1, textor2006volcanic2}. However, the resolution of ATHAM is still pretty coarse compared with SK-3D and PDAC.
Besides strengthening their strength (ATHAM has been adding more and more microphysics, PDAC was extended to consider more phases), these models are learning from each other. PDAC is trying to include effect of microphysics into the model while ATHAM wants to extend its ability to modeling pyroclastic flow. Both are using finer and finer resolution. Recently, a first order, non-equilibrium compressible 3D model, ASHEE \citep{cerminara2016ashee}, is derived based on three dimensional N phases Eulerian transportation equation. ASHEE is valid for low concentration and low Strokes number region and much faster than N-phase Eulerian model. The model is based an open source numerical solver OpenFOAM adopting unstructured Finite Volume method.\\
To summarize, each 3D model has its own focus based on problem of interest. Different simplification was made based on their scope. Improvements have been making since their birth. Accuracy (depends on comprehensiveness of the model, resolution of discretization, numerical error, and order of accuracy) and simulation time (depends on number of governing equations, resolution, numerical methods and parallel techniques) are always a couple of conflict considerations in 3D plume simulations.\\

@@

All of these existing 3D models are based on mesh based methods (Eulerian methods). And there is no 3D plume model based on mesh free method (Lagrange methods) yet. Actually, Lagrangian methods have several features that are suitable for volcano plume simulation. Among such Lagrangian methods, Smoothed Particle Hydrodynamics \citep{gingold1977smoothed,lucy1977numerical}, even though relatively newer compared to well designed mesh based methods such as Finite Difference, has shown good agreement with experiments by many applications in fluid dynamic area. And it currently is, by far the most widely used scheme. Implementations of SPH in volcanology are keeping showing up \citep{bursik2003smoothed,herault2010sph,haddad2016smoothed}.\\
Specifically, the reasons that motivate us to choose SPH as the numerical method for volcanic plume simulation are as following:

@@

\item Investigation of mixing phenomena. 
\item ZFE (Zone of Flow Establishment), ZEF (Zone of Established Flow) investigation and relation to column collapse and the mystery of the development of entrainment and proper characterization through numerical experimentation.
\item Related to below, easy inclusion of particles of different sizes (phases) and investigation of detailed mechanics of sedimentation and drag force interaction in lower plume.
\item For mesh based method, either interface tracking (Lagrangian) \citep{harlow1965numerical, wrobel1991computational, cheng1995simplified} or interface capturing (Eulerian) \citep{hirt1981volume, youngs1982time, gerlach2006comparison, gopala2008volume} methods are used to reconstruct the flow interface of free boundary flow. High computational cost, a tendency to form numerical instabilities and the inability to track complex topological changes are the significant drawbacks of Lagrangian techniques \citep{hirt1981volume, unverdi1992front, anderson1998diffuse}. For interface capturing (Eulerian) method, the surrendering of surface detail before the phase transport calculation means that interface reconstruction is required between time steps to recover the interface information, which need additional numerical effort \citep{hirt1981volume, youngs1982time}. Being able to adaptively adjust the discretization and automatically construct the interface, SPH does not require additional numerical effort for interface construction and therefor more suitable for volcano plume simulation.

@@

\item With more than one material, each described by its own set of particles, interface problems between phases are often trivial for SPH but difficult for mesh based schemes. So multiphase flow can be easily handled by SPH. Adding of new phases to the model also does not require a major overhaul and re-tooling. As will be shown in later paragraph, adding of new phases will lead to adding of several lines into the source code for new phases and additional interaction terms between existing phases and newly added phases.

@@

As discussed in the previous paragraph, existing 3D plume models focus on one or several specific aspects of plume and have been extended to be more comprehensive by accounting for more physics or more phases. Easy extensibility and capability of handling multiples phase flow without additional numerical effort greatly facilitate future extension of SPH models. As volcanic plume are in nature multiphase and without pre-defined boundary in the atmosphere, SPH is a suitable numerical method for plume modeling. The more basic physics, such as entrainment of air and thermal expansion, are essential for all plume modeling while some other physics, such as water condense and aggregation ect., are important in specific scenario. As an initial effort on exploiting advantages of SPH in volcano plume modeling, we focus on capturing more basic features in plume development. So our current model is far away from comprehensive. Fortunately, the easy extension feature of SPH facilitate developing of a more comprehensive model in the future.\\

@@

\subsection{Our contribution}
As SPH is a relative new method in computational fluid dynamics, implementation of SPH for compressible multiphase turbulent flow is scarce. Several remedies of classical SPH are adopted in our implementation. One of the most popular implementations of SPH is simulating of free surface flow, such as breaking-waves and flood. Less attention was paid to velocity inlet and pressure outlet boundary conditions which are required in plume modeling. We impose pressure boundary condition by adding extra layers of static ghost particles. Additional constrain on time step is used to avoid growing up of numerical fluctuations near the pressure boundary. We impose velocity inlet boundary condition by placing several layers of ghost particles moving with eruption velocity. Turbulence model is crucial for reproducing the entrainment of air. There are several turbulence models proposed for SPH method \citep{issa2005numerical,violeau2007numerical}. We adopt a LANS (Lagrangian Averaged Navier-Stokes) turbulence model \citep{monaghan2011turbulence}, which was originally proposed for incompressible, and extend it for compressible flow accounting for turbulent heat exchange. Corrected formulism of SPH is adopted to bypass the well-known tensile instability issues of classical SPH.\\

@@

The pressure of the surrounding atmosphere is given. Except for the pressure, boundary conditions for density, velocity, and energy on the outlet should depend on the solution. As we ignore the viscosity, the shear stress is gone and norm stress (whose magnitude equals to pressure) will balance with ambient pressure.

@@

Ignore derivative terms higher than first order, and write the intergral in particle approximation form:

@@

\section{Code availability}

>>>>

%% Copernicus Publications Manuscript Preparation Template for LaTeX Submissions
%% ---------------------------------
%% This template should be used for copernicus.cls
%% The class file and some style files are bundled in the Copernicus Latex Package, which can be downloaded from the different journal webpages.
%% For further assistance please contact Copernicus Publications at: production@copernicus.org
%% http://publications.copernicus.org/for_authors/manuscript_preparation.html
%% Please use the following documentclass and journal abbreviations for discussion papers and final revised papers.
%% 2-column papers and discussion papers
\documentclass[journal abbreviation, manuscript]{copernicus}
%% Journal abbreviations (please use the same for discussion papers and final revised papers)
% Archives Animal Breeding (aab)
% Atmospheric Chemistry and Physics (acp)
% Advances in Geosciences (adgeo)
% Advances in Statistical Climatology, Meteorology and Oceanography (ascmo)
% Annales Geophysicae (angeo)
% ASTRA Proceedings (ap)
% Atmospheric Measurement Techniques (amt)
% Advances in Radio Science (ars)
% Advances in Science and Research (asr)
% Biogeosciences (bg)
% Climate of the Past (cp)
% Drinking Water Engineering and Science (dwes)
% Earth System Dynamics (esd)
% Earth Surface Dynamics (esurf)
% Earth System Science Data (essd)
% Fossil Record (fr)
% Geographica Helvetica (gh)
% Geoscientific Instrumentation, Methods and Data Systems (gi)
% Geoscientific Model Development (gmd)
% Geothermal Energy Science (gtes)
% Hydrology and Earth System Sciences (hess)
% History of Geo- and Space Sciences (hgss)
% Journal of Sensors and Sensor Systems (jsss)
% Mechanical Sciences (ms)
% Natural Hazards and Earth System Sciences (nhess)
% Nonlinear Processes in Geophysics (npg)
% Ocean Science (os)
% Proceedings of the International Association of Hydrological Sciences (piahs)
% Primate Biology (pb)
% Scientific Drilling (sd)
% SOIL (soil)
% Solid Earth (se)
% The Cryosphere (tc)
% Web Ecology (we)
% Wind Energy Science (wes)
%% \usepackage commands included in the copernicus.cls:
%\usepackage[german, english]{babel}
%\usepackage{tabularx}
%\usepackage{cancel}
%\usepackage{multirow}
%\usepackage{supertabular}
%\usepackage{algorithmic}
%\usepackage{algorithm}
%\usepackage{amsthm}
%\usepackage{float}
%\usepackage{subfig}
%\usepackage{rotating}
\begin{document}
%<<<<<<< HEAD

\title{Plume-SPH 1.0: A 3D dust-gas volcano plume model based on Smoothed Particle Hydrodynamics}

%=======
\title{Plume-SPH 1.0: A 3D dust-gas volcano plume model based on smoothed particle hydrodynamics}
%>>>>>>> 0d96d9682fb484b8b159e8d09a0eb89723d3a0f8
% \Author[affil]{given_name}{surname}
\Author[1]{Zhixuan}{Cao}
\Author[1]{Abani}{Patra}
\Author[2]{Marcus}{Bursik}
\Author[3]{E. Bruce}{Pitman}
\Author[4]{Matthew}{Jones}
\affil[1]{Department of Mechanical and Aerospace Engineering, University at Buffalo, SUNY, New York, United State}
\affil[2]{Department of Geology, University at Buffalo, SUNY, New York, United State}
\affil[3]{Department of Material Design and Innovation, University at Buffalo, SUNY, New York, United State}
\affil[4]{Center for Computational Research,
University at Buffalo, SUNY, New York, United State}
%% The [] brackets identify the author with the corresponding affiliation. 1, 2, 3, etc. should be inserted.
\runningtitle{Plume-SPH 1.0}
\runningauthor{Z. Cao et al}
%<<<<<<< HEAD

\correspondence{Abani Patra (abani@buffalo.com)}

%=======
%\correspondence{Abani Patra (abani.patra@gmail.com)}
%>>>>>>> 0d96d9682fb484b8b159e8d09a0eb89723d3a0f8
\received{}
\pubdiscuss{} %% only important for two-stage journals
\revised{}
\accepted{}
\published{}
%% These dates will be inserted by Copernicus Publications during the typesetting process.
\firstpage{1}
\maketitle
\begin{abstract}
%Volcanic ash, if it persists in the atmosphere, will pose great threat to aviation and communities. Therefore, accurate modeling and forecasts are crucial. 
VATDs (Volcanic ash transport and dispersion models) used for critical volcanic ash forecasts usually take outputs of eruption plume models as source terms. The accuracy of source terms is crucial for forecasts from such VATDs. All  existing 3D (three dimensional) plume models described in the literature use mesh based methods. SPH (Smoothed particle hydrodynamics), as a mesh free method has several advantages over mesh based methods in modeling of multiphase free boundary flow. As an initial effort on exploiting the feasibility and advantages of SPH in volcano plume modeling we adopt a relatively simpler model, a 3D dusty-gas dynamic model, targeted at capturing the salient features of volcano plume. In this model, erupted material is assumed to be well mixed and represented by one phase while air is another phase. We also assume immediate dynamic and thermal dynamic equilibrium between air and erupted material and that the effect of wind is not significant.

Several newly developed techniques in SPH are adopted and extended to address numerical challenges in simulating multiphase compressible turbulent flow with classical SPH method. The $SPH-\varepsilon$ turbulence model is adopted to capture mixing at unresolved scales. Heat exchange due to turbulence is then calculated with Reynolds analogy. Corrected formalism of SPH is used to handle tensile instability and deficiency of particle distribution near the boundaries. We also propose feasible ways to impose velocity inlet and pressure outlet boundary conditions, both of which are scarce in traditional implementations of SPH. The core solver of our model is parallelized with MPI (message passing interface) obtaining good weak scalability and strong scalability. The model is first verified by comparing velocity and concentration distribution along axis and cross section with experimental results of JPUE (jet or plume which is ejected from a nozzle into a uniform environment). The top height of Pinatubo eruption (15 June 1991) simulated by our model is consistent with both observation and existing 3D plume models. Profiles of several integrated variables are compared with these by existing 3D plume models and further verifies the correctness of our model. Analysis on the plume evolution process illustrated that this model is able to reproduce the basic physics of plume development. The comparison also implies that turbulence model plays a significant role in 3D volcano plume modeling.
\end{abstract}
\introduction  %% \introduction[modified heading if necessary]
%<<<<<<< HEAD
\subsection{Volcanic ash hazards}

%The dynamics of explosive eruption clouds has been a central issue in volcanology  for a long time \citep{settle1978volcanic, wilson1978control}.
 Primary hazards associated with explosive grey volcanoes include pyroclastic density currents (flows and surges), the widespread deposition of airfall tephra, and the threats to aviation posed by volcanic ash in the upper atmosphere
% In addition, there are also other secondary hazards associated with volcanoes 
\citep{lockwood2013volcanoes}.
Simulation of all possible hazards with one model is difficult due to diversity among these hazards. Our focus here are  the risks that volcanic ash posed on aircrafts and communities. 
During volcanic eruptions, VATDs (volcanic ash transport and dispersion models) are used to forecast the location and movement of ash clouds over hours to days.  VATDs use eruption source parameters, such as plume height, mass eruption rate, duration, and the mass fraction distribution of erupted debris finer than about $4 \Phi$ (or $63  \mu$m), which can remain in the cloud for many hours or days. Observational data for such parameters are usually unavailable in the first minutes or hours after an eruption is detected. Moreover, these input parameters are subject to change during an eruption, requiring rapid re-assignment of new parameters. Usually, plume models are used to provide these source terms for VATDs and the forecast accuracy is critically dependent on these models. This paper reports on a new 3D (three dimensional) volcano plume model designed to exploit the advantages of mesh-free methods for 3D modeling of such plumes  that involve multi-phase free boundary flows.\\
%=======
The dynamics of explosive eruption clouds has been a central issue of volcanology science for a long time \citep{settle1978volcanic, wilson1978control}.
The basic physics of volcanic plumes has been studied for more than half a century \citep{morton1956turbulent, settle1978volcanic, wilson1978control}. The effects of magma type and vent conditions \citep{woods1988fluid, woods1995decompression}, atmospheric conditions \citep{ woods1993moist, sparks1997volcanic, bursik2001effect}, external surface water \citep{koyaguchi1996formation}, thermal disequilibrium, and particle fallout \citep{woods1991particle} have been assessed in increasing details based on observation, experiments and numerical studies.

@@

%Hazards associated with explosive grey volcanoes include pyroclastic density currents (flows and surges), the widespread deposition of airfall tephra, and the threats posed by volcanic ash. 

@@

%>>>>>>> 0d96d9682fb484b8b159e8d09a0eb89723d3a0f8

@@

%Simulation of all possible hazards with one model is difficult due to diversity among these hazards. Our concern in current work is the risks that volcanic ash poses on aircrafts and communities. %During volcanic eruptions, VATDs (volcanic ash transport and dispersion models) are used to forecast the location and movement of ash clouds over hours to days. Those VATDs use eruption source parameters, such as plume height, mass eruption rate, duration, and the mass fraction distribution of erupted debris finer than about $4 \Phi$ (or $63  \mu$m), which can remain in the cloud for many hours or days. Observational data for such parameters are usually unavailable in the first minutes or hours after an eruption is detected. Moreover, these input parameters are subject to change during an eruption, requiring rapid re-assignment of new parameters. Usually, plume models are used to provide source terms for VATDs. 

@@

Several 1D (one dimensional) volcano plume models have been developed in the past few decades. Ranging from the most basic 1D model \citep{woods1988fluid} which only accounts for mass conservation to more recently developed 1D models (eg. \citep{bursik2001effect, mastin2007user, degruyter2012improving, woodhouse2013interaction, devenish2013using, de2015plume, folch2016fplume, pouget2016sensitivity}) which tend to account for more complex physics effects. FPLUME-1.0 \citep{folch2016fplume} accounts for wind effect, entrainment of moisture, water phase change, particle fallout and re-entrainment and even wet aggregation of ash. However, in these 1D models, the entrainment of air is evaluated based on two coefficients: entrainment coefficient for the vertical plume and the entrainment coefficient that describes the effect of wind. Different 1D models adopt different entrainment coefficients based on specific formulation or calibration against well-documented case studies. The feedback from plume to atmosphere is also ignored in 1D models. Even though determination of essential parameters such as the entrainment is not based on first principles, such simple models nevertheless allow us to investigate the importance of physical mechanisms in a volcanic plume. In addition, these simplified models require little computational resource and can run on standard personal computers or on web sites in very short time. As a result, 1D software for volcano plume development (such as \citep{267, 1194, 3541}), combined with VATDs (such as \citep{114, draxler2015hysplit}) are widely used in research and practice. While these 1D models can generate well-matched results with 3D (three dimensional) models for weak plumes, much greater variability is observed for strong plume scenarios, especially for local variables \citep{costa2016results}. In addition, the need for greater skill in hazards forecasts especially where the plume model is used to generate source conditions for complex VATD models requires 2D (two dimensional) and 3D models.

The development of 2D and 3D, time-dependent, and multiphase numerical models for volcanic plumes has provided new explanations for many features of explosive volcanism. One of the earliest of these is the 3D model PDAC (Pyroclastic Dispersion Analysis Code) \citep{neri2003multiparticle}  which is a non-equilibrium, multiphase, 3D compressible flow model. Conservation equations for each phase are solved separately with finite volume method.  A parallel computing version of PDAC was developed lately \citep{ongaro2007parallel}. Advanced numerical techniques, such as second order scheme and semi-explicit time stepping, was also adopted afterwards to improve the accuracy of PDAC \citep{carcano2013semi}. 
Another 3D model, SK-3D \citep{suzuki2005numerical} is a 3D time-dependent fluid dynamics model that attempts to reproduce the entrainment process of eruption clouds with relatively simpler physics but with high order numerical accuracy and high spatial resolution.
%which is applicable to time-dependent phenomena in actual volcanological simulations. 
A series of simulations based on SK-3D was reported, including determination of the entrainment coefficients of eruption columns as a function of height \citep{suzuki2010numerical},  the effect of wind field on entrainment coefficient \citep{suzuki20133d}, establishment of the relationship between the observable quantities of the eruption clouds and the eruption conditions at the vent \citep{suzuki2009three} (previously, such relationship is based on 1D models and greatly depends on empirical constants), investigation of the effect of the intensity of turbulence in the umbrella cloud on dispersion and sedimentation of tephra \citep{koyaguchi2009effect}. 
While the SK-3D focuses on accurately capturing the entrainment caused by turbulent mixture with higher resolution and numerical method of higher order, PDAC takes the disequilibrium between different phases into account and hence is a true multiphase model. Another 3D model, ATHAM ( Active Tracer High-Resolution Atmospheric Model) \citep{oberhuber1998volcanic} focuses more on microphysics within the plume. As pyroclastic flow is not the initial concern of ATHAM, immediately dynamics and thermodynamics equilibrium is assumed in ATHAM. The dynamic core of ATHAM solves the compressible Euler equations for momentum, pressure and temperature of the gas particle mixture \citep{oberhuber1998volcanic}. The subgrid-scale turbulence closure scheme differentiates between the horizontal and vertical directions \citep{herzog2003prognostic} to capture turbulent mixing. The cloud microphysics predicts the mass of hydrometeors in liquid and ice phase \citep{herzog1998effect}. Additional modules include radiation \citep{langmann1997radiative}, gas phase chemistry \citep{trentmann2002simulation} and gas scavenging by hydrometeors \citep{textor2003injection} were added lately. A further extension was made to include particle aggregation in 2006 \citep{textor2006volcanic1, textor2006volcanic2}. However, the resolution of ATHAM is still pretty coarse compared with SK-3D and PDAC.
Besides adding to their special strengths (ATHAM has been adding more and more microphysics, PDAC was extended to consider more phases), these models are learning from each other. PDAC developers have moved to include the effect of microphysics into the model while ATHAM developer have extended its ability to modeling pyroclastic flow. Both are using finer and finer resolution. Recently, a first order, non-equilibrium compressible 3D model, ASHEE \citep{cerminara2016ashee}, was introduced based on three dimensional N phases Eulerian transportation equation. ASHEE is valid for low concentration and low Stokes number region and much faster than N-phase Eulerian model. The model is based an open source numerical solver OpenFOAM adapting its unstructured Finite Volume solver.\\
To summarize, each 3D model has its own focus based on problem of interest and modeling/numerical choices made.
 %Different simplification was made based on their scope. Improvements have been making since their birth. 
 Accuracy of simulation and forecast skill (depends on comprehensiveness of the model, resolution of discretization, numerical error, and order of accuracy) and simulation time (depends on number of governing equations, resolution, numerical methods and parallel techniques) .%are always a couple of conflict considerations in 3D plume simulations.\\

@@

All of these existing 3D models are mesh based Eulerian methods. To date there are no 3D plume models based on mesh free Lagrangian methods. Lagrangian methods have several features that are suitable for volcano plume simulation that we outline below. Among such Lagrangian methods, Smoothed Particle Hydrodynamics \citep{gingold1977smoothed,lucy1977numerical}, %even though relatively newer compared to well designed mesh based methods such as Finite Difference,
 based simulations have shown good agreement with experiments for many applications in fluid dynamics. And it currently is, by far the most widely used meshfree scheme. Implementations of %SPH in volcanology  may be seen in \citep{bursik2003smoothed,herault2010sph,haddad2016smoothed}.\\
Specifically, we choose SPH as the numerical method for volcanic plume simulation to enable:

@@

\item better investigation of mixing phenomena;
\item accurate modeling of the development of ZFE (Zone of Flow Establishment), ZEF (Zone of Established Flow) investigation and relation to column collapse and the questions relating to the development of entrainment; %and proper characterization through numerical experimentation.
and,
\item  easy inclusion of particles of different sizes (phases) and investigation of detailed mechanics of sedimentation and drag force interaction in lower plume.
These  will be enabled by the following features of SPH:
\item For mesh based method, either interface tracking (Lagrangian) \citep{harlow1965numerical, wrobel1991computational, cheng1995simplified} or interface capturing (Eulerian) \citep{hirt1981volume, youngs1982time, gerlach2006comparison, gopala2008volume} methods are used to reconstruct the flow interface of free boundary flow. High computational cost, a tendency to form numerical instabilities and the inability to track complex topological changes are the significant drawbacks of tracking techniques \citep{hirt1981volume, unverdi1992front, anderson1998diffuse}. For interface capturing (Eulerian) method, the surrendering of surface detail before the phase transport calculation means that interface reconstruction is required between time steps to recover the interface information, which needs additional numerical effort \citep{hirt1981volume, youngs1982time}. Since SPH is able to adaptively adjust the discretization and automatically construct the interface, SPH does not require additional numerical effort for interface construction and therefore is more suitable for volcano plume simulation.

@@

\item With more than one material, each described by its own set of particles, interface problems between phases are often trivial for SPH but difficult for mesh based schemes. So multiphase flow can be easily handled by SPH. Adding of new phases to the model also does not require a major overhaul and re-tooling. As will be shown in later paragraphs, adding of new phases will lead to adding of several lines into the source code for new phases and additional interaction terms between existing phases and newly added phases.

@@

As discussed in the previous paragraph, existing 3D plume models focus on one or several specific aspects of plume and have been extended to be more comprehensive by accounting for more physics or more phases. Easy extensibility and capability of handling multiples phase flow without additional numerical effort greatly facilitate future extension of SPH models. As volcanic plume are in nature multiphase and without pre-defined boundary in the atmosphere, SPH is a suitable numerical method for plume modeling. The core physics, such as entrainment of air and thermal expansion, are essential for all plume modeling while some other physics, such as water condensation and aggregation etc., are important in specific scenario. As an initial effort on exploiting advantages of SPH in volcano plume modeling, we focus on capturing basic features in plume development using a robust numerically and comuptationaly efficient framework with support for scalable parallel computing. %So our current model is far away from comprehensive. 
Open source availability and  the easy extension feature of SPH will facilitate developing of a more comprehensive community driven model.\\

@@

\subsection{Our contributions}
As SPH is a relatively new method in computational fluid dynamics, implementations of SPH for compressible multiphase turbulent flows are few. Several issues endemic to classical SPH necessary for use in this domain are fixed in our implementation. The most popular applications of SPH (and their original motivating application) has been in simulating of free surface flow, such as breaking-waves and flood. Less attention was paid to velocity inlet and pressure outlet boundary conditions which are required in plume modeling. 
\item We develop methodology to impose pressure boundary conditions by adding extra layers of static ghost particles. Additional constraints on time step is used to avoid the growth of numerical fluctuations near the pressure boundary. We impose velocity inlet boundary condition by placing several layers of ghost particles moving with eruption velocity. 
\item Turbulence model is crucial for reproducing the entrainment of air. There are several turbulence models proposed for SPH method \citep{issa2005numerical,violeau2007numerical}. We adopt a LANS (Lagrangian Averaged Navier-Stokes) turbulence model \citep{monaghan2011turbulence}, which was originally proposed for incompressible, and is extended here for compressible flow accounting for turbulent heat exchange. 
\item Corrected formulation of SPH \citep{kumar2013parallel} is adapted to bypass the well-known tensile instability issues of classical SPH.\\

@@

The pressure of the surrounding atmosphere is given. Except for the pressure, boundary values for density, velocity, and energy on the outlet should depend on the solution. As we ignore the viscosity, the shear stress is ignored and normal stress (whose magnitude equals to pressure) will balance the ambient pressure.

@@

Ignoring derivative terms higher than first order, and writing the integral in particle approximation form:

@@

\subsection{Code availability}
